---
title: "Stroke Prediction"
author: "Cynthia Matuli"
date: "2026-05-02"
output:
  pdf_document: default
  html_document: default
---
 


      
      
  **PREDICTIVE MODELLING FOR STROKE**


Stroke is a leading cause of disability and death worldwide with LMICs accounting for three-quarters of all stroke cases. These are, however, deaths and complications that are preventable with early detection. Data Science methods such as Machine Learning predictive modelling and classification techniques including logistic regression, Gradient Boosting Trees, and Random Forest have stronger generalization and feature expression capabilities to learn and map complex nonlinear relationship between variables; and therefore, better at predicting high risk stroke patients.


- Primary objective is to develop a predictive model for stroke.


- Secondary objectives;

  - Compare the predictive performance of:
  
    - Logistic regression
    
    - Decision tree
    
    - Random forest
    
    - XGBoost
    
  - Select the best-performing model.
  
  - Report the important predictors from that model.
  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  message = FALSE, warning = FALSE)

```



```{r results = 'hide', Libraries}

library(writexl); library(openxlsx); library(tinytex); library(formatR)
library(readxl); library(xgboost); library(caret); library(Matrix)
library(tidyverse); library(forcats); library(stringr); library(nnet)
library(readxl);library(dplyr);library(ggplot2);library(pROC);library(ggthemes);
library(gtsummary);library(flextable);library(broom.helpers); library(xfun);
library(corrplot); library(rsample); library(recipes); library(parsnip); library(tune)
library(dials); library(yardstick); library(naniar); library(baguette)

```



```{r Dataset}

stroke_prediction <- read_excel("ANALYSIS DATASETS/stroke_prediction.xlsx")
View(stroke_prediction)


#Structure of the dataset
str(stroke_prediction)

```



**Understanding the dataset:

  - The dataset has 5110 entries and 12 variables. 3 numerical variables and 9 categorical variables.
  
  - id: unique identifier
  
  - gender: "Male", "Female" or "Other"
  
  - age: age of the patient
  
  - hypertension: 0 if the patient does not have hypertension, 1 if the patient has hypertension
  
  - heart_disease: 0 if the patient does not have any heart diseases, 1 if the patient has a heart disease.
  
  - ever_married: "No" or "Yes".
  
  - work_type: "children", "Govt_jov", "Never_worked", "Private" or "Self-employed".
  
  - Residence_type: "Rural" or "Urban".
  
  - avg_glucose_level: average glucose level in blood.
  
  - bmi: body mass index
  
  - smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown".
  
  - stroke: 1 if the patient had a stroke or 0 if not.
  



```{r Glimpse of the dataset}

#Overview of the unique entries of the dataset
glimpse(stroke_prediction)


#Summary of the variables
stroke_prediction %>%
summary()

```


**Analysis of the unique entries:

  - bmi is read as a character yet it is a numerical variable.
  
  - hypertension, heart_disease, and stroke - should be converted to factor variables.
  
  - Age has a minimum of 0.08. Very unlikely especially for an individual likely to suffer from stroke.
  
  


```{r Data completeness and duplicates}

#Looking for duplicates
stroke_prediction %>% 
  count(id) %>% 
  arrange(desc(n))


#Missing values and the nature of missingness
stroke_prediction %>% 
  naniar::vis_miss()

```



  - The dataset has no duplicates.
  
  - There are no missing values. Further exploration of the unique entries through descriptive statistics 




**Variable preprocessing**

```{r Stroke}

#Recoding stroke to read names instead of numbers
table(stroke_prediction$stroke)

stroke_prediction <- stroke_prediction %>% 
  mutate(stroke_recorded = case_when(
    stroke == 0 ~ "No stroke",
    stroke == 1 ~ "Stroke",
    TRUE ~ NA
  )) %>% 
  mutate(stroke_recorded = factor(stroke_recorded,
                                  levels = c("No stroke", "Stroke"))) %>% 
  relocate(stroke_recorded, .after = stroke)
  
table(stroke_prediction$stroke_recorded)

```



```{r Gender}

#Removing 'Other' in gender since it is only 1 entry
table(stroke_prediction$gender)

stroke_prediction <- stroke_prediction %>% 
  filter(!(gender == "Other")) %>% 
  mutate(gender = factor(gender,
                         levels = c("Female", "Male")))

table(stroke_prediction$gender)

```



```{r All ages histogram}

#Recoding age into groups.
summary(stroke_prediction$age) #Minimum age is 0.08 and max is 82.0
stroke_prediction$age <- as.numeric(stroke_prediction$age)

stroke_prediction <- stroke_prediction %>% 
  mutate(age = floor(age)) %>% #Rounding down age to remove decimals
  arrange(age)


#Histogram to explore age
stroke_prediction %>% 
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black",
                 boundary = min(stroke_prediction$age, na.rm = TRUE)) +
  stat_bin(binwidth = 10, 
           geom = "text", 
           aes(label = after_stat(count)), 
           vjust = -0.5, 
           boundary = min(stroke_prediction$age, 
                          na.rm = TRUE)) +
  labs(x = "Age (years)", y = "Age Count") +
  theme_minimal()

```



Majority of the individuals are between 50-60 years.


There are some children in the dataset and predicting their stroke chances will most likely skew or mislead the model. I will structure the dataset into several parts.
  
  1. Create a dataset for adults 40 years of age and above (>= 40yrs)
  
  2. Create a dataset for individuals >=18years of age (I'll work with this later)
  


```{r Age histogram}

#Eliminating individuals below 40years of age
stroke_prediction <- stroke_prediction %>% 
  filter(age >= 40)

age_histo <- stroke_prediction %>% 
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black",
                 boundary = min(stroke_prediction$age, na.rm = TRUE)) +
  stat_bin(binwidth = 10, 
           geom = "text", 
           aes(label = after_stat(count)), 
           vjust = -0.5, 
           boundary = min(stroke_prediction$age, 
                          na.rm = TRUE)) +
  labs(x = "Age (years)", y = "Count") +
  theme_minimal()

ggsave(filename = "Distribution of age.png", plot = age_histo, width = 6, height = 4, dpi = 600)

```



```{r Age groups}

#Recoding age into age-groups
stroke_prediction <- stroke_prediction %>% 
  mutate(age_group = case_when(
    age < 45 ~ "Young adults (<45)",
    age >= 45 & age < 65 ~ "Middle aged (45-64)",
    age >= 65 ~ "Elderly (>=65)",
    TRUE ~ NA
  )) %>% 
  mutate(age_group = factor(age_group,
                            levels = c("Young adults (<45)", "Middle aged (45-64)", "Elderly (>=65)"))) %>% 
  relocate(age_group, .after = age)

table(stroke_prediction$stroke_recorded, stroke_prediction$age_group)

```


```{r Work type}

#Exploring work type.
table(stroke_prediction$work_type)

```


Does not show evidence of data sparsity.




```{r bmi histogram}

#Converting bmi to numeric
class(stroke_prediction$bmi)
stroke_prediction$bmi <- as.numeric(stroke_prediction$bmi)
summary(stroke_prediction$bmi)



#Visualizing the distribution of bmi
bmi_histo <- stroke_prediction %>% 
  ggplot(aes(x = bmi)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black",
                 boundary = min(stroke_prediction$bmi, na.rm = TRUE)) +
  stat_bin(binwidth = 10, 
           geom = "text", 
           aes(label = after_stat(count)), 
           vjust = -0.5, 
           boundary = min(stroke_prediction$bmi, 
                          na.rm = TRUE)) +
  labs(x = "bmi", y = "Count") +
  theme_minimal()

ggsave(filename = "Distribution of bmi.png", plot = bmi_histo, width = 6, height = 4, dpi = 600)

```



- Summary of bmi: its shows 145 missing entries. I shall impute when doing feature engineering to prevent data leakage. I shall impute with the median value.

- The histogram shows that bmi is skewed to the right.




```{r bmi categories}

#Recoding bmi levels into clinical ranges
stroke_prediction <- stroke_prediction %>% 
  mutate(bmi_weight_obesity = case_when(
    bmi < 18.5 ~ "Underweight (<18.5)",
    bmi >= 18.5 & bmi < 25.0 ~ "Normal (18.5-24.9)",
    bmi >= 25.0 & bmi < 30.0 ~ "Overweight (25.0-30.0)",
    bmi >= 30.0 ~ "Obesity (>=30.0)",
    TRUE ~ NA
  )) %>% 
  mutate(bmi_weight_obesity = factor(bmi_weight_obesity,
                                     levels = c("Obesity (>=30.0)", "Underweight (<18.5)", "Normal (18.5-24.9)", "Overweight (25.0-30.0)"))) %>% 
  relocate(bmi_weight_obesity, .after = bmi)

table(stroke_prediction$bmi_weight_obesity)

table(stroke_prediction$bmi_weight_obesity, stroke_prediction$stroke_recorded)

```


- Reference for the clinical ranges: https://apps.who.int/nutrition/landscape/help.aspx?menu=0&helpid=420



The recoded bmi variable based on the clinical ranges shows a high degree of data sparsity. Individuals recoded as underweight are only 25 and only 1 individual has stroke. This would lead to unstable model estimates. I will collapse this group and combine it with individuals recoded as having normal weight.



```{r bmi categories collapsed}

#Recoding bmi levels into clinical ranges
stroke_prediction <- stroke_prediction %>% 
  mutate(bmi_weight_obesity = case_when(
    bmi < 25.0 ~ "Underweight/Normal (<25.0)",
    bmi >= 25.0 & bmi < 30.0 ~ "Overweight (25.0-30.0)",
    bmi >= 30.0 ~ "Obesity (>=30.0)",
    TRUE ~ NA
  )) %>% 
  mutate(bmi_weight_obesity = factor(bmi_weight_obesity,
                                     levels = c("Obesity (>=30.0)", "Underweight/Normal (<25.0)", "Overweight (25.0-30.0)"))) %>% 
  relocate(bmi_weight_obesity, .after = bmi)

table(stroke_prediction$bmi_weight_obesity)

table(stroke_prediction$bmi_weight_obesity, stroke_prediction$stroke_recorded)

```


```{r Glucose level histogram}

#Visualizing the distribution of average glucose level
summary(stroke_prediction$avg_glucose_level)

glucose_histo <- stroke_prediction %>% 
  ggplot(aes(x = avg_glucose_level)) +
  geom_histogram(binwidth = 15, fill = "steelblue", color = "black",
                 boundary = min(stroke_prediction$avg_glucose_level, na.rm = TRUE)) +
  stat_bin(binwidth = 15, 
           geom = "text", 
           aes(label = after_stat(count)), 
           vjust = -0.5, 
           boundary = min(stroke_prediction$avg_glucose_level, 
                          na.rm = TRUE)) +
  labs(x = "Average Glucose Level", y = "Count") +
  theme_minimal()

ggsave(filename = "Distribution of glucoselevels.png", plot = glucose_histo, width = 6, height = 4, dpi = 600)

```



The distribution is bimodal.




```{r Glucose level}

#Recoding average glucose levels into clinical ranges
stroke_prediction <- stroke_prediction %>% 
  mutate(avg_glucose_level_diabetes = case_when(
    avg_glucose_level < 117 ~ "Normal (<117mg/dL)",
    avg_glucose_level >= 117 & avg_glucose_level < 137 ~ "Prediabetes (117-136mg/dL)",
    avg_glucose_level >= 137 ~ "Diabetes (>=137mg/dL)",
    TRUE ~ NA
  )) %>% 
  mutate(avg_glucose_level_diabetes = factor(avg_glucose_level_diabetes,
                                             levels = c("Normal (<117mg/dL)", "Prediabetes (117-136mg/dL)", "Diabetes (>=137mg/dL)"))) %>% 
  relocate(avg_glucose_level_diabetes, .after = avg_glucose_level)

table(stroke_prediction$avg_glucose_level_diabetes)

```


- Reference for the clinical ranges: https://my.clevelandclinic.org/health/diagnostics/9731-a1c


- The groups seems okay.




```{r Smoking status}

#Recoding smoking status to smokers and non-smokers and unknown (is quite a %)
table(stroke_prediction$smoking_status)

#Combining formerly smoked and smokes into 1 group
stroke_prediction <- stroke_prediction %>% 
  mutate(smoking_status_recorded = case_when(
    smoking_status == "never smoked" ~ "Non-smokers",
    smoking_status %in% c("formerly smoked", "smokes") ~ "Smokers",
    smoking_status == "Unknown" ~ "Unknown"
  )) %>% 
  mutate(smoking_status_recorded = factor(smoking_status_recorded,
                                          levels = c("Non-smokers", "Smokers", "Unknown"))) %>% 
  relocate(smoking_status_recorded, .after = smoking_status)

table(stroke_prediction$smoking_status_recorded)

```



```{r Ever married}

#ever_married into a factor
table(stroke_prediction$ever_married)

stroke_prediction <- stroke_prediction %>% 
  mutate(ever_married = factor(ever_married))


```


- 91% of the individuals have ever been married.




```{r Heart disease}

#heart_disease
table(stroke_prediction$heart_disease)

stroke_prediction <- stroke_prediction %>% 
  mutate(heart_disease_recorded = case_when(
    heart_disease == 0 ~ "No",
    heart_disease == 1 ~ "Yes",
    TRUE ~ NA)) %>% 
  mutate(heart_disease_recorded = factor(heart_disease_recorded,
                                         levels = c("No", "Yes"))) %>% 
  relocate(heart_disease_recorded, .after = heart_disease)

table(stroke_prediction$heart_disease_recorded)

```


- 90% of the individuals do not have heart disease




```{r Hypertension}

#hypertension
table(stroke_prediction$hypertension)

stroke_prediction <- stroke_prediction %>% 
  mutate(hypertension_recorded = case_when(
    hypertension == 0 ~ "No",
    hypertension == 1 ~ "Yes",
    TRUE ~ NA)) %>% 
  mutate(hypertension_recorded = factor(hypertension_recorded,
                                         levels = c("No", "Yes"))) %>% 
  relocate(hypertension_recorded, .after = hypertension)

table(stroke_prediction$hypertension_recorded)


```


- 84% of the individuals do not have hypertension.




**Assessing the association between variables and the outcome of interest**


*Hypothesis testing*

  1. Chi-square test
  
    - Testing whether categorical variables have any association with the outcome variable (stroke)
    
    - Hypothesis;
    Null: Predictor variables have no association with the dependent variable.
    Alternative: Predictor variables have an association with the dependent variable.
    
    - Significance level; 0.05
    
    
    
  2. Wilcoxon test
  
    - Testing whether there is a difference in the medians of numerical variables.
    
    - Hypothesis;
    Null: There is no difference in the median values between groups.
    Alternative: There is a difference in the median value between groups.
    
    - Significance level; 0.05
    

```{r Descriptive statistics}

#Descriptive statistics
descriptive <- stroke_prediction %>% 
  select(gender:age_group, hypertension_recorded, heart_disease_recorded, ever_married:bmi, bmi_weight_obesity, smoking_status_recorded, stroke_recorded) %>% 
  tbl_summary(by = stroke_recorded,
              type = list(all_categorical() ~ "categorical"), missing = "ifany") %>% 
  add_overall() %>% 
  add_p()
descriptive
```



- Gender, ever married, work type, residence type, bmi categorical, and smoking status have an insignificant association with stroke (the chi-square p-values are greater than 0.05) while age groups, hypertension, heart disease, and glucose level categorical have a significant association with stroke (chi-square p-values less than 0.05).


- The median age of participants is 58 years with individuals with stroke having a median age of 72 and those without stroke having a median age of 57.


- 41% of the individuals are smokers.


- 9% have a heart disease


- 16% have hypertension.


- The median bmi level is 30 across both stroke and non-stroke individuals. 


- The median average glucose level is 95. Individuals with stroke have a median level 107 while those without stroke have a median level of 94. Individuals with stroke have significantly higher glucose levels compared to individuals without stroke.





```{r Correlation}

#Correlation between the continuous variables
correlation <- stroke_prediction %>% 
  select(age, bmi, avg_glucose_level) %>% 
  cor(use = "complete.obs", method = "pearson")
correlation

correlationplot <- corrplot(
  correlation,
  method = "color",
  type = "upper", 
  addCoef.col = "black",     # Adding correlation coefficients
  tl.col = "black",          # Variable label color
  tl.srt = 45               # Rotating labels
)


png(filename = "Correlation of variables.png", width = 6, height = 4, units = "in",
  res = 600)

corrplot(correlation, method = "color", type = "upper", addCoef.col = "black",
  tl.col = "black", tl.srt = 45)

dev.off()

```




- bmi, glucose level, and age have very weak relationships with each other.


  - Age and bmi have a very weak negative relationship.
  

  - Age and glucose level and bmi and glucose level have a very weak positive relationship. 





```{r Data completeness}

#Missingness
missing <- stroke_prediction %>% 
  miss_var_summary()
missing


#Visualize
stroke_prediction %>% 
  arrange(desc(age)) %>% 
  naniar::vis_miss()

```



- bmi has barely 5% missing values.





**Train-Test Split**


```{r Train Test data split}

#Setting random seed for reproducibility
set.seed(123)

#Splitting the dataset
dataset_split <- initial_split(stroke_prediction, prop = 0.70, strata = stroke_recorded)

training_data <- dataset_split %>% 
  training()

testing_data <- dataset_split %>% 
  testing()

```




*Determining the proportion of stroke in training, testing sets and the whole dataset*


```{r Data split proportions of Stroke}

#WHOLE DATASET
#Proportions/distributions in the overall/main dataset
table(stroke_prediction$stroke_recorded)


#TRAINING SET
table(training_data$stroke_recorded)
#Percentages for both "No stroke" and "Stroke"
prop.table(table(training_data$stroke_recorded))


#TESTING SET
table(testing_data$stroke_recorded)
#Percentages for both "No stroke" and "Stroke"
prop.table(table(testing_data$stroke_recorded))

```



**Feature Engineering**


```{r Feature Engineering Post data split}
#Selecting the variables I need for modelling.

colnames(stroke_prediction)

training_data <- training_data%>%
  select(gender, age_group, hypertension_recorded, heart_disease_recorded, ever_married, work_type, Residence_type, avg_glucose_level_diabetes, bmi, bmi_weight_obesity, smoking_status_recorded, stroke_recorded)

View(training_data)

```





```{r Feature engineering using recipe}
library(themis)

#bmi values are Missing Completely At Random (MCAR), therefore, it is safe to do median imputation


#Creating a recipe and performing upsampling to deal with class imbalance
stroke_recipe <- recipe(stroke_recorded ~ .,
                        data = training_data) %>% 
  step_impute_median(bmi)%>% #Median imputation
  step_string2factor(all_categorical()) %>% #Converting categorical vars to factors
  step_mutate(
    bmi_weight_obesity = case_when(
    bmi < 25.0 ~ "Underweight/Normal (<25.0)",
    bmi >= 25.0 & bmi < 30.0 ~ "Overweight (25.0-30.0)",
    bmi >= 30.0 ~ "Obesity (>=30.0)",
    TRUE ~ NA
  )) %>% 
  step_mutate(bmi_weight_obesity = factor(bmi_weight_obesity,
                                     levels = c("Obesity (>=30.0)",
                                                "Underweight/Normal (<25.0)",
                                                "Overweight (25.0-30.0)")))%>%
  step_rm(bmi) %>%
  step_upsample(stroke_recorded)
  
#Upsampling performed better than SMOTE, so I moved forward with upsampling.

```


- I have performed upsampling on the outcome variable to prevent the model from making predictions based on the majority class only.




**LOGISTIC REGRESSION MODEL**


```{r Logistic Regression Model}

#Defining the model
stroke_model <- logistic_reg() %>% 
  set_engine('glm') %>% 
  set_mode('classification')

stroke_model


library(workflows)
#Creating a workflow
stroke_workflow <- workflows::workflow()%>%
  add_model(stroke_model)%>%
  add_recipe(stroke_recipe)

stroke_workflow


#Fitting/Training the model
stroke_workflow_fit <- stroke_workflow %>% 
  fit(data = training_data)


```


  *Testing set Predictions*

```{r Obtaining Logistic model predictions}

#Obtaining model predictions
logistic_class <- predict(stroke_workflow_fit,
                          new_data = testing_data,
                          type = "class")

logistic_prob <- predict(stroke_workflow_fit,
                         new_data = testing_data,
                         type = "prob")

logistic_predictions <- bind_cols(logistic_class,
                                  logistic_prob,
                                  testing_data)%>% 
  relocate(stroke_recorded, .before = .pred_class)


View(logistic_predictions)

```




  *Confusion matrix*

```{r Logistics Confusion matrix}

#Confusion matrix
conf_mat(logistic_predictions, truth = stroke_recorded, estimate = .pred_class)


conf_mat(logistic_predictions,
         truth = stroke_recorded,
         estimate = .pred_class)%>%
  summary()

```



- Accuracy is not the best metric to consider because our classes are highly imbalanced. In this case **Sensitivity** (capturing true positives), **balanced accuracy** (overall performance accounting for class imbalance), **ROC AUC** (overall model discrimination), **PR AUC {Precision-Recall AUC}** (minority class performance, **F1-score** (trade-off between precision and recall) give better model evaluation)





  *Model performance*

```{r Logistics Model performance}

library(yardstick)

logistic_model_metrics <- metric_set(yardstick::sensitivity,
                                      yardstick::bal_accuracy,
                                      yardstick::precision,
                                      yardstick::recall,
                                      yardstick::f_meas,
                                      yardstick::specificity)

logistic_model_metrics(logistic_predictions,
               truth = stroke_recorded,
               estimate = .pred_class,
               event_level = "second") #Stroke is the event of interest


```



```{r Evaluating model performance via ROC AUC and PR AUC}

#ROC plot
roc_logit <- logistic_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second")

roc_logit %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_path(linewidth = 1) +
  geom_abline(linetype = 2, colour = "blue") +
  coord_equal()

#ROC-AUC value
logistic_predictions %>%
  roc_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")


#PR AUC
logistic_predictions %>%
  pr_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")

```


```{r Yeah Obtaining fitted model coefficients}

library(broom)
library(dplyr)
library(gt)

results_table <- extract_fit_parsnip(stroke_workflow_fit) %>%
  tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(`Odds Ratio` = round(estimate, 2),
         `95% CI` = paste0(round(conf.low, 2)," – ", round(conf.high, 2)),
         `P-value` = if_else(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
         Significant = if_else(p.value < 0.05, "Yes", "No")) %>%
  arrange(p.value)%>%
  rename(Predictor = term) 

results_table %>%
  select(Predictor, `Odds Ratio`, `95% CI`, `P-value`, Significant) %>%
  gt() %>%
  cols_label(Predictor = "Predictor") %>%
  tab_header(title = md("**Multivariable Logistic Regression Model**"),
             subtitle = "Predictors of Stroke") %>%
  fmt_markdown(columns = everything())


```



- In the upsampled multivariable logistic regression model, age, hypertension, overweight status, heart disease, and diabetes (glucose level >=137) were among the strongest predictors of stroke.


- In the multivariable logistic regression model;


  - Age group was the strongest predictor of stroke.
  
  
    - Compared with young adults, elderly individuals had substantially higher odds of stroke (adjusted OR = 51.63, 95% CI: 26.98–115.01, p < 0.001), while middle-aged adults also had significantly increased odds (adjusted OR = 18.26, 95% CI: 9.56–40.61, p < 0.001).
    
  
  - Individuals with hypertension have 86% increased odds (95% CI: 1.55–2.23, p < 0.001) of having stroke compared to individuals without hypertension.
  
  
  - Individuals with an overweight status have a 61% increased odds (95% CI: 1.28–1.77, p < 0.001) of having stroke compared to individuals who have a bmi >=30.0
  
  
  -Individuals with heart disease 65% increased odds (95% CI: 1.33–2.09, p < 0.001) of having stroke compared to individuals without heart disease.





**DECISION TREE MODEL**


```{r Decision Tree}

#Defining the model
dt_stroke_model <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

dt_stroke_model


#Specifying the workflow
dt_stroke_workflow <- workflow() %>%
  add_model(dt_stroke_model) %>%
  add_recipe(stroke_recipe)

dt_stroke_workflow


#Fitting the model
dt_stroke_model_fit <- dt_stroke_workflow %>%
  fit(data = training_data)




```



```{r Obtaining DT model predictions}

#Model predictions
dt_class <- predict(dt_stroke_model_fit, new_data = testing_data, type = "class")

dt_prob <- predict(dt_stroke_model_fit, new_data = testing_data, type = "prob")

dt_predictions <- bind_cols(dt_class, dt_prob, testing_data) %>%
  relocate(stroke_recorded, .before = .pred_class)

View(dt_predictions)

```




```{r DT Confusion matrix and model performance}

#Confusion matrix
conf_mat(dt_predictions, truth = stroke_recorded, estimate = .pred_class)


accuracy_metrics <- metric_set(
  yardstick::sensitivity, bal_accuracy, precision, recall,f_meas, yardstick::specificity)

accuracy_metrics(dt_predictions, truth = stroke_recorded, estimate = .pred_class,
               event_level = "second") #Stroke is the event of interest


```



```{r DT AUC and ROC plot}

#ROC plot
#ROC plot
roc_dt <- dt_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second")

roc_dt %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_path(linewidth = 1) +
  geom_abline(linetype = 2, colour = "blue") +
  coord_equal()


#ROC-AUC value
dt_predictions %>%
  roc_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")

#PR-AUC value
dt_predictions %>%
  pr_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")

```



```{r Decision Tree Hyperparameter tuning}

dt_model_tuning <- decision_tree(
  cost_complexity = tune(),
  min_n = tune(),
  tree_depth = tune()) %>%
  set_engine("rpart") %>%
  set_mode("classification")

dt_model_tuning



#Workflow
dt_tune_workflow <- workflow() %>%
  add_model(dt_model_tuning) %>%
  add_recipe(stroke_recipe)

dt_tune_workflow


#Creating a regular grid
dt_tune_grid <- grid_regular(extract_parameter_set_dials(dt_model_tuning), levels = 5)


#Cross validation to select the best hyperparameters
set.seed(123)

stroke_cv_folds <- vfold_cv(training_data, v = 10)

dt_tuning_results <- tune_grid(dt_tune_workflow,
                               resamples = stroke_cv_folds,
                               grid = dt_tune_grid,
                               metrics = metric_set(roc_auc))

#Visualization of the tuning results
autoplot(dt_tuning_results)


collect_metrics(dt_tuning_results)


show_best(dt_tuning_results, metric = "roc_auc")


#Optimal hyperparameters
best_params <- select_best(dt_tuning_results, metric = "roc_auc")
best_params


#Finalizing the model and fit to make predictions
best_specs <- finalize_model(dt_model_tuning, best_params)
best_specs


#Generating the final workflow
final_dt_workflow <- dt_tune_workflow %>%
  finalize_workflow(best_params)


#Generating final results using the last_fit() function to train, test, make predictions and test automatically
final_dt_results <- final_dt_workflow %>%
  last_fit(split = dataset_split,
           metrics = metric_set(roc_auc, pr_auc, yardstick::sensitivity, yardstick::specificity, bal_accuracy, f_meas))

dt_tuned_predictions <- collect_predictions(final_dt_results)

collect_metrics(final_dt_results)

```







**RANDOM FOREST MODEL**


```{r Random Forest}
library(ranger)

#Defining the model
rf_stroke_model <- rand_forest() %>%
  set_engine("ranger") %>%
  set_mode("classification")

rf_stroke_model



#Defining a workflow
rf_stroke_workflow <- workflow() %>%
  add_model(rf_stroke_model) %>%
  add_recipe(stroke_recipe)

rf_stroke_workflow


#Model fit
rf_stroke_model_fit <- rf_stroke_workflow %>%
  fit(data = training_data)


```


```{r RF Model predictions}

#Model predictions
rf_class <- predict(rf_stroke_model_fit, new_data = testing_data, type = "class")

rf_prob <- predict(rf_stroke_model_fit, new_data = testing_data, type = "prob")

rf_predictions <- bind_cols(rf_class, rf_prob, testing_data) %>%
  relocate(stroke_recorded, .before = .pred_class)

View(rf_predictions)

```




```{r RF Model Evaluation}

#Confusion matrix
conf_mat(rf_predictions, truth = stroke_recorded, estimate = .pred_class)


accuracy_metrics <- metric_set(yardstick::sensitivity, bal_accuracy, precision, recall,f_meas, yardstick::specificity)

accuracy_metrics(rf_predictions, truth = stroke_recorded, estimate = .pred_class,
               event_level = "second") #Stroke is the event of interest


#ROC plot
roc_rf <- rf_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second")

roc_rf %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_path(linewidth = 1) +
  geom_abline(linetype = 2, colour = "blue") +
  coord_equal()


#ROC-AUC value
rf_predictions %>%
  roc_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")

#PR-AUC value
rf_predictions %>%
  pr_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")

```



```{r RF Hyperparameter Tuning}

rf_model_tuning <- rand_forest(mtry = tune(),
                               min_n = tune(),
                               trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("classification")

rf_model_tuning



#Workflow
rf_tune_workflow <- workflow() %>%
  add_model(rf_model_tuning) %>%
  add_recipe(stroke_recipe)

rf_tune_workflow


#Creating a random grid
rf_params <- extract_parameter_set_dials(rf_tune_workflow) %>%
  finalize(training_data)

rf_tune_grid <- grid_random(rf_params, size = 10) #10 random hyperparameter combinations


#Cross validation to select the best hyperparameters
set.seed(123)

stroke_cv_folds <- vfold_cv(training_data, v = 10)

rf_tuning_results <- tune_grid(rf_tune_workflow,
                               resamples = stroke_cv_folds,
                               grid = rf_tune_grid,
                               metrics = metric_set(roc_auc))


collect_metrics(rf_tuning_results)


show_best(rf_tuning_results, metric = "roc_auc")


#Optimal hyperparameters
best_params <- select_best(rf_tuning_results, metric = "roc_auc")
best_params


#Finalizing the model and fit to make predictions
best_specs <- finalize_model(rf_model_tuning, best_params)
best_specs


#Generating the final workflow
final_rf_workflow <- rf_tune_workflow %>%
  finalize_workflow(best_params)


#Generating final results using the last_fit() function to train, test, make predictions and test automatically
final_rf_results <- final_rf_workflow %>%
  last_fit(split = dataset_split,
           metrics = metric_set(roc_auc, pr_auc, yardstick::sensitivity, yardstick::specificity, bal_accuracy, f_meas))

rf_tuned_predictions <- collect_predictions(final_rf_results)

collect_metrics(final_rf_results)

```





**XGBOOST MODEL**


```{r XGBOOST}

#Defining the model
xgboost_model <- boost_tree() %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgboost_model



#Defining the recipe
xgboost_recipe <- stroke_recipe %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_zv(all_predictors())



#Defining the workflow
xgboost_workflow <- workflow() %>%
  add_model(xgboost_model) %>%
  add_recipe(xgboost_recipe)

xgboost_workflow



#Fitting the model
xgboost_workflow_fit <- xgboost_workflow %>%
  fit(data = training_data)



```


```{r XGBOOST Model predictions}

#Model predictions
xgb_class <- predict(xgboost_workflow_fit, new_data = testing_data, type = "class")

xgb_prob <- predict(xgboost_workflow_fit, new_data = testing_data, type = "prob")

xgboost_predictions <- bind_cols(xgb_class, xgb_prob, testing_data) %>%
  relocate(stroke_recorded, .before = .pred_class)

View(xgboost_predictions)

```



```{r XGBOOST Model evaluation}

#Confusion matrix
conf_mat(xgboost_predictions, truth = stroke_recorded, estimate = .pred_class)


accuracy_metrics <- metric_set(yardstick::sensitivity, bal_accuracy, precision, recall,f_meas, yardstick::specificity)

accuracy_metrics(xgboost_predictions, truth = stroke_recorded, estimate = .pred_class,
               event_level = "second") #Stroke is the event of interest


#ROC plot
roc_xgb <- xgboost_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second")

roc_xgb %>%
  ggplot(aes(x = 1 - specificity,
             y = sensitivity)) +
  geom_path(linewidth = 1) +
  geom_abline(linetype = 2, colour = "blue") +
  coord_equal()


#ROC-AUC value
xgboost_predictions %>%
  roc_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")

#PR-AUC value
xgboost_predictions %>%
  pr_auc(truth = stroke_recorded, .pred_Stroke, event_level = "second")

```



```{r XGBOOST Hyperparameter Tuning}

xgb_model_tuning <- boost_tree(mtry = tune(),
                               min_n = tune(),
                               trees = tune(),
                               tree_depth = tune(),
                               learn_rate = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgb_model_tuning



#Workflow
xgb_tune_workflow <- workflow() %>%
  add_model(xgb_model_tuning) %>%
  add_recipe(xgboost_recipe)

xgb_tune_workflow


#Creating a random grid
xgb_params <- extract_parameter_set_dials(xgb_tune_workflow) %>%
  finalize(training_data)

xgb_tune_grid <- grid_random(xgb_params, size = 10) #10 random hyperparameter combinations


#Cross validation to select the best hyperparameters
set.seed(123)

stroke_cv_folds <- vfold_cv(training_data, v = 10)

xgb_tuning_results <- tune_grid(xgb_tune_workflow,
                               resamples = stroke_cv_folds,
                               grid = xgb_tune_grid,
                               metrics = metric_set(roc_auc))


collect_metrics(xgb_tuning_results)


show_best(xgb_tuning_results, metric = "roc_auc")


#Optimal hyperparameters
best_params <- select_best(xgb_tuning_results, metric = "roc_auc")
best_params


#Finalizing the model and fit to make predictions
best_specs <- finalize_model(xgb_model_tuning, best_params)
best_specs


#Generating the final workflow
final_xgb_workflow <- xgb_tune_workflow %>%
  finalize_workflow(best_params)


#Generating final results using the last_fit() function to train, test, make predictions and test automatically
final_xgb_results <- final_xgb_workflow %>%
  last_fit(split = dataset_split,
           metrics = metric_set(roc_auc, pr_auc, yardstick::sensitivity, yardstick::specificity, bal_accuracy, f_meas))

collect_metrics(final_xgb_results)

xgb_predictions <- collect_predictions(final_xgb_results)

head(xgb_predictions)

```




**COMPARING THE MODELS PERFORMANCE**


```{r Comparing models}
library(yardstick)

comparison_metrics <- metric_set(roc_auc, pr_auc, yardstick::sensitivity, yardstick::specificity, precision, bal_accuracy, f_meas)


#Logistic Regression
logistic_results <- comparison_metrics(logistic_predictions,
                                       truth = stroke_recorded,
                                       estimate = .pred_class,
                                       .pred_Stroke,
                                       event_level = "second") %>%
  select(.metric, .estimate) %>%
  rename(Logistic_Regression = .estimate)



#Tuned Decision Tree
dt_results <- comparison_metrics(dt_predictions,
                                 truth = stroke_recorded,
                                 estimate = .pred_class,
                                 .pred_Stroke,
                                 event_level = "second") %>%
  select(.metric, .estimate) %>%
  rename(Decision_Tree = .estimate)



#Tuned Random Forest
rf_results <- comparison_metrics(rf_tuned_predictions,
                                 truth = stroke_recorded,
                                 estimate = .pred_class,
                                 .pred_Stroke,
                                 event_level = "second") %>%
  select(.metric, .estimate) %>%
  rename(Tuned_Random_Forest = .estimate)



#Tuned XGBOOST model
xgb_results <- comparison_metrics(xgb_predictions,
                                  truth = stroke_recorded,
                                  estimate = .pred_class,
                                  .pred_Stroke,
                                  event_level = "second") %>%
  select(.metric, .estimate) %>%
  rename(Tuned_XGBoost = .estimate)


#Combing them
models_comparison <- logistic_results %>%
  left_join(dt_results, by = ".metric") %>%
  left_join(rf_results, by = ".metric") %>%
  left_join(xgb_results, by = ".metric")

#Comparison table
models_comparison %>%
  mutate(across(-.metric, ~ round(.x, 3))) %>%
  gt() %>%
  cols_label(
    .metric = "Performance Metric",
    Logistic_Regression = "Logistic\nRegression",
    Decision_Tree = "Decision\nTree",
    Tuned_Random_Forest = "Random\nForest",
    Tuned_XGBoost = "XGBoost"
  ) %>%
  tab_header(
    title = md("**Comparison of Final Stroke Prediction Models**"),
    subtitle = "Performance based on the Unseen (independent) Testing Set"
  )

```




```{r ROC AUC models comparison}

# Logistic Regression ROC
roc_logistic <- logistic_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second") %>%
  mutate(Model = "Logistic Regression (AUC = 0.733)")


# Decision Tree ROC
roc_dt <- dt_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second") %>%
  mutate(Model = "Decision Tree (AUC = 0.706)")


# Random Forest ROC
roc_rf <- rf_tuned_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second") %>%
  mutate(Model = "Random Forest (AUC = 0.711)")


# XGBOOST ROC
roc_xgb <- xgb_predictions %>%
  roc_curve(truth = stroke_recorded, .pred_Stroke, event_level = "second") %>%
  mutate(Model = "XGBOOST (AUC = 0.742)")


# Combine all ROC curves
roc_all <- bind_rows(roc_logistic, roc_dt, roc_rf, roc_xgb)


# Plotting ROC curves
roc_plot <- ggplot(roc_all, aes(x = 1 - specificity, y = sensitivity, colour = Model)) +
  geom_path(linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  coord_equal() +
  labs(
    title = "Receiver Operating Characteristic (ROC) Curves",
    subtitle = "Comparison of Final Stroke Prediction Models",
    x = "False Positive Rate (1 − Specificity)",
    y = "True Positive Rate (Sensitivity)",
    colour = "Prediction Model"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"))

ggsave(filename = "ROC.png", plot = roc_plot, width = 6, height = 4, dpi = 600)

```



*Model Performance Interpretations*


- I compared four machine learning models/algorithms on the stroke prediction testing set.


  - XGBoost model demonstrated the best overall predictive performance, achieving the highest ROC AUC (0.742) with a sensitivity of 0.708. The model also recorded the highest Precision-Recall AUC (0.205).
  
  
  - Logistic Regression model showed competitive performance with an AUC value of 0.733 and attaining the highest sensitivity of 0.736. Although the XGBOOST model is better at discriminating between individuals with stroke and those without stroke, the logistic regression model captured more individuals with stroke compared to XGBOOST.
  
  
  - The decision tree had the lowest AUC (0.706), however, it recorded the highest sensitivity (0.736) as the logistic regression model. Further, the decision tree has higher precision, balanced accuracy, F1-score compared to every other model. 
  
  
- Generally, the best performing model was **XGBOOST**, followed by **Decision Tree**, **Logistic Regression**, and finally **Random Forest**.


- While the models were able to identify a good percentage of stroke cases, the precision was fairly low accounting for the very low number of stroke cases (high class imbalance).
  
  




```{r XGBOOST Feature Importance}

xgb_fit <- final_xgb_results %>%
  extract_workflow() %>%
  extract_fit_parsnip()


#Feature Importance Table
importance_tbl <- vip::vi(xgb_fit)

importance_tbl_clean <- importance_tbl %>%
  mutate(Variable = case_when(
      Variable == "age_group_Elderly" ~ "Elderly (≥65)",
      Variable == "age_group_Middle.aged" ~ "Middle aged (45–64)",
      Variable == "avg_glucose_level" ~ "Average glucose level",
      Variable == "avg_glucose_level_diabetes_Diabetes....137mg.dL." ~
        "Diabetes (≥137mg/dL)",
      Variable == "bmi_weight_obesity_Overweight..25.0.30.0." ~
        "Overweight (25.0–30.0)",
      Variable == "bmi_weight_obesity_Underweight.Normal...25.0." ~
        "Underweight/Normal (<25.0)",
      Variable == "hypertension_recorded_Yes" ~ "Hypertension",
      Variable == "heart_disease_recorded_Yes" ~ "Heart disease",
      Variable == "smoking_status_recorded_Smokers" ~ "Smokers",
      Variable == "smoking_status_recorded_Unknown" ~
        "Smoking status unknown",
      Variable == "gender_Male" ~ "Male",
      Variable == "Residence_type_Urban" ~ "Urban residence",
      Variable == "ever_married_Yes" ~ "Ever married",
      Variable == "work_type_Private" ~ "Private employment",
      Variable == "work_type_Self.employed" ~ "Self-employed",
      TRUE ~ Variable)) %>%
  arrange(desc(Importance))

importance_tbl_clean



#Feature Importance Plot

#Method 1
#vip::vip(xgb_fit, num_features = nrow(importance_tbl), geom = "col")


#Method 2
library(ggplot2)
library(forcats)

vip_plot <- importance_tbl_clean %>%
  ggplot(aes(x = Importance, y = fct_reorder(Variable, Importance))) +
  geom_col() +
  labs(
    title = "Feature Importance from Tuned XGBoost Model",
    x = "Relative Importance (Gain)",
    y = "Predictor") +
  theme_minimal()


ggsave(filename = "Feature Importance.png", plot = vip_plot, width = 6, height = 4, dpi = 600)

```



*Interpretation*


- **Highly influential predictors:** Elderly age group (>=65 years), middle-aged group (45-65 years), average glucose level (Diabetic (>=137mg/dL)), heart disease, and hypertension.


- **Lower-contributing predictors:** Smoking status, work type, gender, residence, marital status, etc.



